Think deep for reef resilience

### 2025-07-02 paper/paper_draft.qmd
Add info on carbonate budgets and start of DHW timeseries plotting

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and coral thermal response and adaptation research into coral population modelling allows us to investigate how these abiotic and biotic factors interact and will influence coral climate change resilience into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015].

It is hypothesised that sink reefs are more likely to be resilient under climate change, compared to source reefs, due to greater influx of coral recruits post disturbance. It is hypothesised that reefs with greater depths are more likely to be resilient under future thermal stress, compared to reefs at shallow depths, as they benefit from thermal refugia. It is hypothesised that reef cover will decline by greater magnitude as the Equilibrium Climate Sensitivity (ECS) of each Global Climate Model (GCM) increases.

@@

@@

## Carbonate budget analysis

Scleractinian corals are primary habitat structures on coral reefs. The accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes such as coral accumulation *Perry*, bioerosion *Tribollet*, physical damage generating coral rubble *Kenyon*, and calcificationa and *CaCO3* dissolution rates *Eyre 2014, 2018*. The balance of these processes and occurrance of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity *Perry, Desbiens*, rates of physical environmental processes and ocean acidity *Perry, Eyre*. It has been suggested that total live coral cover on a reef highly influences the maintainence of net positive carbonate budgets. Estimates of live coral cover required to maintain net positive carbonate budget levels range from 10 to 20% *Perry* of a reef's total area, with a level of 17% live coral cover for the Great Barrier Reef *Desbiens*. Using median absolute cover timeseries for each reef we calculated the number of years each reef maintains coral cover above these live coral cover threshold levels, inferring the number of years each reef maintains a net positive carbonate budget, accumulating reef framework. As in situ carbonate budgets are influenced by a range of processes, our calculations of carbonate budget maintainence are intended as estimates and do not reflect the conditions that will be experienced on a reef, but indicate relative levels of health across reefs. To capture the wide influence of environmental factors on carbonate budget levels, we analyse live coral cover thresholds required for positive carbonate budgets from 10 to 20% (inclusive, at step 1%).

@@

Timeseries analysis has been used to identify three groups of reefs with similar declining (low), recovering (high), and intermediate (medium) coral cover trajectories throughout future scenarios. Under all GCMs most reefs suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. Differences between low, medium and high reef clusters in their timeseries are most notable under these two GCMs.

@@

<!-- liu2024 citation -->

>>>>

Scleractinian corals are primary habitat structures on coral reefs. The accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes such as coral accumulation *Perry*, bioerosion *Tribollet*, physical damage generating coral rubble *Kenyon*, and calcificationa and *CaCO3* dissolution rates *Eyre 2014, 2018*. The balance of these processes and occurrance of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity *Perry, Desbiens*, rates of physical environmental processes and ocean acidity *Perry, Eyre*. It has been suggested that total live coral cover on a reef highly influences the maintainence of net positive carbonate budgets. Estimates of live coral cover required to maintain net positive carbonate budget levels range from 10 to 20% *Perry* of a reef's total area, with a level of 17% live coral cover for mid and outershelf reefs on the Great Barrier Reef *Desbiens*.

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and coral thermal response and adaptation research into coral population modelling allows us to investigate how abiotic and biotic factors interact and will influence coral climate change resilience and the maintainence of positive carbonate budgets into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015].

It is hypothesised that sink reefs are more likely to be resilient under climate change, compared to source reefs, due to greater influx of coral recruits post disturbance. We predict that increasing connectivity levels will correlate with increasing maintainence of positive carbonate budgets. It is hypothesised that reefs with greater depths are more likely to be resilient under future thermal stress, compared to reefs at shallow depths, as they benefit from thermal refugia. We predict that increasing reef depth will correlate with increasing maintainence of positive carbonate budgets. It is hypothesised that reef cover will decline by greater magnitude as the Equilibrium Climate Sensitivity (ECS) of each Global Climate Model (GCM) increases.

@@

### Carbonate budget analysis

Using median coral cover timeseries for each reef we calculated the number of years each reef maintains coral cover above these live coral cover threshold levels, inferring the number of years each reef maintains a net positive carbonate budget, accumulating reef framework. As in situ carbonate budgets are influenced by a range of processes, our calculations of carbonate budget maintainence are intended as estimates and do not reflect the conditions that will be experienced on a reef, but indicate relative levels of health across reefs. To capture the wide influence of environmental factors on carbonate budget levels, we analyse live coral cover thresholds required for positive carbonate budgets from 10 to 20% (inclusive, at step 1%).

@@

@@

Under all GCMs most reefs suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. Differences between low, medium and high reef clusters in cover timeseries are most notable under these two GCMs.

@@
